Investigating Determinism of Physics Engines for usage in CAV and V\&V simulations

### 2021-04-19 document/source/body.tex
rename top level files, fixed oreilly ref error
added a-star section

\subsection{Actor Navigation}
There is existing evidence to suggest actor navigation, typically pedestrians, could be the cause of non-deterministic simulation behaviour~\cite{CARLABenchmark}. Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation~\cite{a_Star_oreilly}. The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle, UnrealAIDocumentation}. A navigation mesh is a fixed area of the environment where actors are free to navigate. Navigation meshes are used by the path planning algorithm to find the shortest distance between navigable points in the environment and could be considered as a potential source of non-determinism if the navigation mesh is not a fixed entity. However, the environment management in CARLA implements fixed binary files for navigation meshes that are linked to each road scene or driving environment. These cannot be unknowingly modified and therefore can be considered fixed.

Unless other sources that can alter the fixed environment exist, the A* algorithm should give deterministic results. Thus, actor navigation should not be considered a source of non-determinism.

\subsection{Summary}
We have investigated the potential sources of non-determinism affecting game engines and explored the impact they may have on simulation variance. Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected. To ensure precise simulation outcomes the physics setting, $dt$, must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters. Non-uniform memory access (NUMA) should only affect interprocessor data transfer cost and, without control measures, will only make the computation cycle longer. Relative access times between different caches are likely to be small although may have a more pronounced impact on high throughput systems, e.g.\ HPC. If this change in computational cycle gives opportunity for the execution order to be changed then this situation may lead to non-determinism.

>>>>

\subsection{Actor Navigation}
There is existing evidence to suggest actor navigation, typically pedestrians, could be a source of non-deterministic simulation behaviour~\cite{CARLABenchmark}. Unreal Engine is the underlying framework for the CARLA simulator which uses the A* search algorithm to find optimal paths for actor navigation~\cite{newton2016unreal}. The optimality and validity of A* solutions is only guaranteed if the environment is deterministic, and if not, then there are no guarantees that an execution of the same A* solution will lead to the same outcome across multiple runs.

To briefly describe the A* process, at each stage of the search, the algorithm selects one node for expansion from a set of nodes known as the \textit{frontier} (also fringe or open set) using a heuristic to select this node. For reasons of efficiency, A* represents the frontier as a priority queue, with nodes prioritised by the heuristic. Selection then reduces to a dequeue operation on the frontier. In general the heuristic does not guarantee a uniquely optimal node in the frontier and thus ties may be encountered. The question of how these ties are broken in a priority queue is related to the broader notion of sorting stability~\cite{Sedgewick2011}. An implementation of a priority queue is said to be stable if it preserves insertion order, otherwise it is said to be unstable. If an implementation of A* uses a stable priority queue, then it will behave as a deterministic algorithm which always returns the optimal node that was first added to the frontier. Stable priority queues are typically less computationally efficient than unstable priority queues, so the requirement on stability is often dropped in practice.

For example, one of the most common implementations of a priority queue relies on a binary heap and is unstable; it will always return the first optimal node according to its internal heap order rather than insertion order. The default priority queue implementations in the standard Java and Boost libraries are based on binary heaps~\cite{boost2021,priorityQueue}, while the Unreal documentation and source code suggests that this is also true of CARLA~\cite{FGraphAStar,TArray,GraphAStar}. Thus, an \textbf{unstable but deterministic} priority queue is an unstable priority queue that preserves some order, but not necessarily insertion order. If an implementation of A* uses an unstable but deterministic priority queue, then it will behave as a deterministic algorithm which always returns the optimal node according the ordering of its priority queue.

Finally, an unstable but deterministic priority queue suggests a third kind of priority queue that we can refer to as an unstable and non-deterministic priority queue; that is, a priority queue that always returns an optimal node selected at random (i.e. breaking ties arbitrarily). If an implementation of A* uses an unstable and non-deterministic priority queue, then it will behave as a non-deterministic algorithm. Overall, an implementation of A* is much more likely to use a priority queue that is unstable but deterministic rather than one that is stable, and is very unlikely to use one that is unstable and non-deterministic.

An implementation of A* will behave as a deterministic algorithm if it uses a stable priority queue or an unstable but deterministic priority queue. However, apparent determinism of A* will be sensitive to the ordering preserved by the priority queue (i.e. insertion order or e.g. heap order). If we are not aware of changes to this ordering, then a deterministic implementation of A* may appear to behave as a non-deterministic algorithm, since we will not have accounted for changes to the ordering. For example, suppose actions are iterated over in order of hash value (where order of iteration over actions often determines insertion order), but between runs there is some untracked change to the software/hardware stack that changes the calculation of the hash value, then this may change insertion order of nodes to the frontier, which could eventually materialise as a different optimal solution found by A*. This suggests that while a simulated environment may behave deterministically, factors outside the simulator may cause changes to the operation of an A* implementation, which would then materialise as non-deterministic runs in the simulator. The difference between the use of a stable priority queue or an unstable but deterministic priority is simply that it may be easier to detect such changes under a stable priority queue because insertion order is typically more meaningful with respect to the actual implementation of A*.

% The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle, UnrealAIDocumentation}. A navigation mesh is a fixed area of the environment where actors are free to navigate. Navigation meshes are used by the path planning algorithm to find the shortest distance between navigable points in then environment and could be considered as a potential source of non-determinism if the navigation mesh is not a fixed entity. However, the environment management in CARLA implements fixed binary files for navigation meshes that are linked to each road scene or driving environment. These cannot be unknowingly modified and therefore can be considered fixed.

% Unless other sources that can alter the fixed environment exist, the A* algorithm should give deterministic results. Thus, actor navigation should not be considered a source of non-determinism.

\subsection{Summary}
We have investigated the potential sources of non-determinism affecting game engines and explored the impact they may have on simulation variance. Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected. To ensure precise simulation outcomes the physics setting, $dt$, must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters. %
%
Any implementation of the A* search algorithm for actor navigation must use a stable priority queue to ensure deterministic results. 
%
Non-uniform memory access (NUMA) should only affect interprocessor data transfer cost and, without control measures, will only make the computation cycle longer. Relative access times between different caches are likely to be small although may have a more pronounced impact on high throughput systems, e.g.\ HPC. If this change in computational cycle gives opportunity for the execution order to be changed then this situation may lead to non-determinism.
